---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# bluewhaledata

Provides sample data for learning how to use the [rstickleback](github.com/FlukeAndFeather/rstickleback) package. 

```{r setup}
# Start by loading a few packages, including rstickleback
library(fs)
library(lubridate)
library(rstickleback)
library(tidyverse)
```

## What's in the data?

[bluewhaledata.zip](bluewhaledata.zip) contains two directories, sensors and events, with CSV files containing bio-logging sensor data and behavioral event timestamps, respectively. For more details about how the data were collected and what they represent, see `[rstickleback::load_lunges()](flukeandfeather.github.io/rstickleback/reference/load_lunges.html)`. 

You can download the data directly in R.

```{r download}
# Create a temporary directory
temp_dir <- tempdir()

# Download file and unzip data
url <- "https://github.com/FlukeAndFeather/bluewhaledata/raw/master/bluewhaledata.zip"
download.file(url, file.path(temp_dir, "bluewhaledata.zip"))
unzip(file.path(temp_dir, "bluewhaledata.zip"), exdir = temp_dir)
```

Each CSV file is named according to the deployment identifier and data type (event or sensor).

```{r dir_struct}
data_dir <- file.path(temp_dir, "data")
# Note: your root folder won't match exactly. It's a random path, generated by
# tempdir().
dir_tree(data_dir)
```

Inspecting the event and sensor data for deployment bw180904-44, we see that event CSV files have one datetime column, `lunge`, and sensor CSV files have five columns. In the sensor data, the `timestamp_utc` datetime column is the time and the other four columns (`depth`, `pitch`, `roll`, and `speed`) describe the animal's movement.

```{r inspect}
sample_events <- read_csv(file.path(data_dir, "events", "bw180904-44_events.csv"))
glimpse(sample_events)
sample_sensors <- read_csv(file.path(data_dir, "sensors", "bw180904-44_sensors.csv"))
glimpse(sample_sensors)
```

To prepare these data for use in rstickleback, we have to combine all the CSVs into two data frames; one each for events and sensors. Here's one way of doing that using a for-loop, which is likely familiar to most readers. I'll demonstrate a more elegant solution later.

```{r read_combine}
# A list of the deployment ids
deployids = c("bw180904-44",
              "bw180904-48",
              "bw180904-52",
              "bw180905-42",
              "bw180905-49",
              "bw180905-53")

# Create empty data frames to hold the results
bw_events_df <- data.frame()
bw_sensors_df <- data.frame()

# Iterate through the deployment ids, read the data, and append to the results
for (id in deployids) {
  # Event data first...
  event_path <- dir(file.path(data_dir, "events"), 
                    pattern = id, 
                    full.names = TRUE)
  events <- read_csv(event_path, col_types = "T") %>% 
    # I happen to know dates are in UTC. Metadata are important!
    mutate(lunge = force_tz(lunge, "UTC"),
           deployid = id)
  bw_events_df <- rbind(bw_events_df, events)
  # ...then sensor data
  sensor_path <- dir(file.path(data_dir, "sensors"), 
                     pattern = id, 
                     full.names = TRUE)
  sensors <- read_csv(sensor_path, col_types = "Tdddd") %>% 
    mutate(timestamp_utc = force_tz(timestamp_utc, "UTC"),
           deployid = id)
  bw_sensors_df <- rbind(bw_sensors_df, sensors)
}

# The results
glimpse(bw_events_df)
glimpse(bw_sensors_df)
```

We now have two data frames containing all the event and sensor data for these six deployments, which we can use to create `Event` and `Sensor` objects for use with `rstickleback`.

```{r events_sensors}
bw_events <- Events(bw_events_df, 
                    deployid_col = "deployid", 
                    datetime_col = "lunge")
bw_sensors <- Sensors(bw_sensors_df,
                      deployid_col = "deployid",
                      datetime_col = "timestamp_utc", 
                      sensor_cols = c("depth", "pitch", "roll", "speed"))

bw_events
bw_sensors
```

These data are now ready to use with `rstickleback` functions. For example, you can visualize the event and sensor data with `[rstickleback::sb_plot_data()](flukeandfeather.github.io/rstickleback/reference/sb_plot.html)` and train a model using `[rstickleback::sb_fit()](flukeandfeather.github.io/rstickleback/reference/sb_fit.html)`.

As a final note, here's the more elegant way to read an entire directory of CSVs using regular expressions to extract the deployment identifier. Note: this only works if your version of `readr` is at least 2.0.0.

```{r read_combine2}
event_csvs <- dir(file.path(data_dir, "events"), full.names = TRUE)
bw_events_df2 <- read_csv(event_csvs, col_types = "T", id = "file_name") %>% 
  mutate(deployid = str_extract(file_name, "bw[0-9]{6}-[0-9]{2}"))
sensor_csvs <- dir(file.path(data_dir, "sensors"), full.names = TRUE)
bw_sensors_df2 <- read_csv(sensor_csvs, col_types = "T", id = "file_name") %>% 
  mutate(deployid = str_extract(file_name, "bw[0-9]{6}-[0-9]{2}"))
```

These data frames do the trick just as well!

```{r events_sensors2}
bw_events2 <- Events(bw_events_df2, 
                     deployid_col = "deployid", 
                     datetime_col = "lunge")
bw_sensors2 <- Sensors(bw_sensors_df2,
                       deployid_col = "deployid",
                       datetime_col = "timestamp_utc", 
                       sensor_cols = c("depth", "pitch", "roll", "speed"))

bw_events2
bw_sensors2
```
